NLIZE: a Perturbation-Driven Visual Interrogation Tool for Analyzing and Interpreting Natural Language Inference Models

### 2018-03-23 paper/intro.tex

% (What is the high level motivation)
% What is the obsatcle for effectively apply machine learning model
With the recent advances in neural network based model, machine learning has gained unprecedented popularity, and its adaption permeated many fields of studies.
%
However, from researchers to practitioners, one often need to overcome many obstacles during the training, debugging, and tuning processes to unleash the full potential of these models.
%
Interpreting the internal mechanism and analyzing how predictions are made are critical to both the design and deployment of a model.
More importantly, the ability to pinpoint where and how an error is made and come up with hypotheses for the cause of failure is the key to identify the limitation and improve upon existing models.
However, providing meaningful answers to these questions are extremely challenging and often regarded as impossible by many.

Recently, substantial research has been developed to combat the model interpretability challenge. Most notably, the works~\cite{SimonyanVedaldiZisserman2013, ZeilerFergus2014, YosinskiCluneNguyen2015, OlahMordvintsevSchubert2017, LiuShiLi2017, OlahSatyanarayanJohnson2018, BilalJourablooYe2018} on interpreting convolution neural network (CNN) model from both machine learning and visualization community have revealed the enormous potential and instant benefits of model interpretation, yet they also remind us how little we actually know about the inner mechanism of these deep neural networks.
%
Compare to image classification task, natural language processing systems have innate restrictions (e.g., the discrete nature of words) that makes the road to interpretation more treacherous. For example, \emph{feature visualization}~\cite{OlahMordvintsevSchubert2017}, an often deployed technique for illustrating what image features a given part of the network (e.g., neuron, layer, or channel) capture, can not be easily generalized to natural language. Image (pixel values) represents a continouous space, in which human can easily recognize the existance (or the absence) of patterns or shapes. For natural language, even though we can encode a word as a vector~\cite{MikolovSutskeverChen2013, PenningtonSocherManning2014}, the word embedding space is still discretely defined, where the interpolation between two vectors (i.e., two words) do not hold clear meaning.
%
These restrictions call for new avenues for solving the interpretation challenges of natural language models, which motivate the proposed work.

% Beside explore the internal states of the model, understand how different parts of the model work together and how one may affect one another is as or if not more important.
Moreover, despite many recent advances, the majority of the existing techniques study the model as an invariant object, where the model's behaviors are recorded and analyzed in an offline fashion.
%
However, the exploratory nature of the model interpretation analysis often leads to many \emph{what if ...} types of questions. Such as, what if I perturbed the current input? Will the prediction be stable? What if we change one of the critical internal states of the model, how would the modification affect the prediction? What if the current prediction is wrong, how and where should we change the model parameters to produce the correct result? These type of queries form a natural way to gain understanding and come up with hypotheses of the model mechanisms by \emph{interrogating} how the components of a model interact with each other in a dynamic setting. Theoretically, we can code up specialized experiments for each of these scenarios. However, such a process is not only tedious but also ignored the iterative nature of the exploratory analysis. Often, these questions are not pre-determined. Instead, new questions arise as we explore the data and analyze previous observations.

In this work, we aim to provide immediate and informative answers to these \emph{what if} questions by exploiting the expressive power of visualization and flexible integration with the underlying computation model. Instead of viewing the model as an invariant object, we approach the interpretability challenge by studying them in a dynamic environment. By employing a perturbation-driven paradigm, we probe the model and examine how changes in one part of the pipeline (the input, internal states, and output prediction, see Fig.~\ref{fig:modelPipeline}) affect others, which in turn provide a new perspective to attack the model interpretation problem.
%
We implement our vision of the perturbation-driven system around the natural language inference problem~\cite{Parikh2016}.
%
However, the componenets of the visualization and the overall paradigm can be easily extended to other NLP tasks, such as neural machine translation, and paragram summerization, etc.
%
In the simplest form, the inference task ask the question about whether \emph{sentence A} can infer the content of \emph{sentence B} (\emph{entail}), or they contradict with each other (\emph{contradiction}), or they talk about different/unrelated things (\emph{neutral}). Language inference addresses the fundimental challenge of sentence semantic similarity, and is one the core natural language process (NLP) tasks (detials in Section~\ref{sec:languageInference}).
%
One recent advance in NLP is the introduction of attention mechanisms~\cite{VaswaniShazeerParmar2017} (see details in Section~\ref{sec:attention}). Intuitively, the \emph{attention} captures where in the input the model should focus on, which assigns importance to individual word or a pair of word (i.e., the alignment between words in different sentences).
%
There has been many speculation on how attention works in various models. In this work, by utilizing the the proposed system, we explore how changes in sentence input, attention, or prediction affect each other. Such type of exploration is not only essentail to interpret the language inference problem, but also other NLP model utilized the same/similar mechansims.
%
Finally, we also enhanced the standard visual encoding (e.g., as a bipariate graph or as an matrix) of the attention matrix by overlaying sentence linguistic structure to allow grammar guided simplification and the ability to examine how attention corresponds to grammarical structure.
%

The key contributions of the proposed works is summarized as the following:
\begin{itemize}
    \item Help understand how the model fail by examining stability of the prediction and visual summarization of predictions;

@@

    \item Enable the interrogation of the relationship between input sentences, attention, and the prediction by interactively exploring how the change in one place affect other parts of the model.

\end{itemize}
%
% In addition,

>>>>

% (What is the high level motivation)
% What is the obsatcle for effectively apply machine learning model
With the recent advances in neural network based model, machine learning has gained unprecedented popularity, and its adaption permeated many fields of studies.
%
However, from researchers to practitioners, one often need to overcome many obstacles during the training, debugging, and tuning processes to unleash the full potential of these models.
%
Interpreting the internal mechanism and analyzing how predictions are made are critical to both the design and deployment of a model.
More importantly, the ability to pinpoint where and how an error is made and come up with hypotheses for the cause of failure is the key to identify the limitation and improve upon existing models.
However, providing meaningful answers to these questions are incredibly challenging and often regarded as impossible by many.

Recently, sizable research has been developed to combat the model interpretability challenges. Most notably, the works~\cite{SimonyanVedaldiZisserman2013, ZeilerFergus2014, YosinskiCluneNguyen2015, OlahMordvintsevSchubert2017, LiuShiLi2017, OlahSatyanarayanJohnson2018, BilalJourablooYe2018} on interpreting convolution neural network (CNN) model from both machine learning and visualization community have revealed the enormous potential and immediate benefits of model interpretation. At the same time, these works also remind us how little we truly understand about the inner mechanism of these deep neural networks.

Compare to image classification task, natural language processing (NLP) systems have innate restrictions (e.g., the discrete nature of words) that makes the road to interpretation more treacherous. For example, \emph{feature visualization}~\cite{OlahMordvintsevSchubert2017}, an often deployed technique for illustrating what image features a given part of the network (e.g., neuron, layer, or channel) capture, can not be readily generalized to natural language. Image (pixel values) represents a continuous solution space, in which human can quickly recognize the existence (or the absence) of patterns or shapes. For natural language, even though we can encode a word as a vector~\cite{MikolovSutskeverChen2013, PenningtonSocherManning2014}, the word embedding space is still discretely defined, in which the interpolation of two vectors (i.e., two words) do not hold clear meaning.
%
These restrictions call for new avenues for solving the interpretation challenges of natural language models, which motivate the proposed work.

% Beside explore the internal states of the model, understand how different parts of the model work together and how one may affect one another is as or if not more important.
Moreover, despite many recent advances, the majority of the existing techniques study the model as an invariant object, where the model's behaviors are recorded and analyzed in an offline fashion.
%
However, the exploratory nature of the model interpretation analysis often leads to many \emph{``what if ...''} type of questions. Such as, what if we perturbed the current input? Will the prediction be stable? What if we change one of the critical internal states of the model, how would the modification affect the prediction? What if the current prediction is wrong, how and where could we apply minimal change to the model to produce the correct result? And how would it affect the internal state we care about? These type of queries form a natural way to gain understanding and come up with hypotheses of the model mechanisms by \emph{interrogating} how the components of a model interact with each other in a dynamic setting. Hypothetically, we can code up specialized experiments for each of these scenarios. However, such a process is not only tedious but also ignores the iterative nature of the exploratory analysis. Often, these questions are not pre-determined. Instead, new exploration paths arise as we investigate and analyze previous observations.

In this work, we aim to provide immediate and informative answers to these \emph{``what if''} questions by combining the expressive power of visualization and direct online query/optimization of the neural network model. Instead of viewing the model as an invariant object, we approach the interpretability challenge by studying them in a dynamic environment. By employing a perturbation-driven paradigm, we probe the internal states of the model and examine how changes in one part of the pipeline (the input, internal states, and output prediction, see Fig.~\ref{fig:modelPipeline}) affect others, which in turn provides a new perspective to address the model interpretation problem.

We implement our vision of the perturbation-driven exploration in an interactive visualization system for natural language inference models~\cite{Parikh2016}.
However, the components of the visualization and the overall paradigm can be readily extended to other NLP tasks, such as neural machine translation, and text summarization, etc.
%
In the simplest form, the inference task ask the question about whether if we can infer the content of \textbf{sentence B} from \textbf{sentence A} (\emph{entail}), or are they contradict with each other (\emph{contradiction}), or are they talk about different/unrelated things (\emph{neutral}). Language inference addresses the fundamental challenge of measuring sentences' semantic similarity, and is one the core natural language processing tasks (see detail in Section~\ref{sec:languageInference}).

One recent advance in neural natural language process models is the introduction of attention mechanisms~\cite{VaswaniShazeerParmar2017} (Section~\ref{sec:attention}). Intuitively, the \emph{attention} help capture which part of the input the model deemed more important for making the right prediction. The attention is often represented as weights for individual words or pairs of words (i.e., the alignment between words in different sentences).
%
%\shusen{seems repeated some of the previous points}
There have been many speculations on how attention works in various models. By utilizing the proposed tool, the domain experts study how changes in sentence input, attention, or prediction affect each other, which help them develop deeper intuitive and alternative hypotheses. 
%we can perturb the attention and observer the change in the prediction, perturb the sentence input and observe the effects on the attention. And more importantly, we also allow the user to change prediction, where we apply a constrained optimization to find the least amount of parameter update to produce the user assign label. 
%Another word, 
%Such type of exploration is not only essential to interpret the language inference problem, but can also be generalized to other NLP model utilized the attention mechanisms.
Moreover, we enhanced the standard visual encoding (e.g., as a bipartite graph or as a matrix) of the attention matrix by overlaying sentence linguistic structure to allow grammar guided simplification of the visual representation.
%
Finally, as discussed in Section~\ref{sec:caseStudy}, the ability to examine how attention corresponds to the grammatical structure also enables domain experts to speculate the potential benefit of including the linguistic structure in the design of attention component of the model.

In summary, the key contributions of the proposed works is the following:
\begin{itemize}
  % \item Distill the visualization tasks for making sense of the role of attention in natural language inference (section~\ref{sec:task});
    \item Introduce the perturbation-driven exploration paradigm for exploring natural language processing models;

    \item Perform error analysis by examining stability of the predictions under perturbation;

@@

    \item Enable the interrogation of the relationship between input sentences, model internals, and the prediction by interactively exploring how the change in one place affect other parts of the model.
\end{itemize}
%
% In addition,

### 2018-03-23 paper/paper_entailVis.tex
update paper structure
update layout of the paper
fix persistence value bug, need to deep copy the python object

%% Paper title.
\title{
NLIVis: a Perturbation-Driven Visual Interrogation Tool for Analyzing and Interpreting Natural Language Inference Models
%Where and How Do Models Fail? A Visual Exploration Tool for Analyzing and Interpreting Natural Language Inference Models
}
%% This is how authors are specified in the journal style

@@

%% Abstract section.
% What are we doing in this paper
\abstract{
With the recent advances in deep learning, neural network based models have obtained the state-of-the-art performance for many linguistic tasks in natural language processing. The disruptive progress also brings enormous challenges. The opaque nature of neural network model leads to hard to debug system and difficult to interpret mechanisms. In this work, we introduce a visualization system that through a tight yet flexible integration between visualization elements and the underlying model, allow a user to interrogate the model by perturbing the input, the internal state, as well as prediction while observing changes in other parts of the pipeline. We use natural language inference task as an example to illustrate how this system help natural language processing (NLP) researchers assess the potential limitation of an NLP model, probe the inner states of the model for interpreting key mechanism such as attention, and ultimately come up with hypotheses on why predictions fail.
} % end of abstract

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{Natural Language Processing, Interpretable Machine Learning,
Visual Analytics}

@@

\input{task.tex}
\input{pipeline.tex}
\input{perturbation.tex} % how to do the perturbation
\input{attention.tex} % how to encode the information
\input{prediction.tex}

\input{application.tex}
%%%%%%%%%%%%% identify errors %%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%% interpret attention %%%%%%%%%%%%%%%%%%%

\input{discussion.tex} % how to encode the information

>>>>

%% Paper title.
\title{
NLIZE: a Perturbation-Driven Visual Interrogation Tool for Analyzing and Interpreting Natural Language Inference Models
%Where and How Do Models Fail? A Visual Exploration Tool for Analyzing and Interpreting Natural Language Inference Models
}
%% This is how authors are specified in the journal style

@@

%% Abstract section.
% What are we doing in this paper
\abstract{
With the recent advances in deep learning, neural network based models have obtained state-of-the-art performances for many linguistic tasks in natural language processing. The disruptive progress also brings enormous challenges. The opaque nature of neural network model leads to hard to debug system and difficult to interpret mechanisms. In this work, we introduce a visualization system that through a tight yet flexible integration between visualization elements and the underlying model allows a user to interrogate the model by perturbing the input, the internal state, as well as prediction while observing changes in other parts of the pipeline. We use natural language inference problem as an example to illustrate how a perturbation-driven paradigm can help domain experts assess the potential limitation of a model, probe its inner states and interpreting and hypothesizing the fundamental model mechanism such as attention.
} % end of abstract

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{Natural Language Processing, Interpretable Machine Learning, Natural Language Inference, Attention Visualization}

@@

\input{task.tex}
\input{pipeline.tex}
\input{perturbation.tex} % how to do the perturbation
\input{sentence.tex}
\input{attention.tex} % how to encode the information
\input{prediction.tex}

\input{application.tex}

\input{evaluation.tex}
\input{discussion.tex} % how to encode the information
